---
title: "Mini Data Analysis Milestone 2 - vancouver_trees dataset"
author: "Yuwei Yang"
date: "2021/10/14"
output: github_document
---


Please install the following four packages before running the code. The package datateachr contains 7 datasets and we will use one dataset for this analysis. The package tidyverse contains the package dplyr and tidyr, and we will use it to do data wrangling and tidy data. The package ggplot2 is used to do visualization in this analysis. The package lubridate provides tools to parse and manipulate dates. 

```{r, message = FALSE}
library(datateachr) #contain the dataset we will be using
library(tidyverse) 
library(ggplot2)
library(lubridate)
```

For this Mini Data Analysis 2, I will be using the vancouver trees dataset. This analysis contains two parts: 1. I will summarize and plot graph using the vancouver tree dataset bassed on four research questions. 2. Tidy the dataset(I will reshape the vancouver trees dataset). 

# Task 1: Process and summarize the data

### 1.1 Research Questions: 
**The following 4 questions are the goal of this analysis:**

1. *How many trees were planted per neighborhood every year?* 
2. *What's the relationship between the height of the tree and the age of the tree for each neighbourhood? * 
3. *Does the diameter of trees different for each species?*
4. *Does the diameter of trees different in each neighbourhood* 

### 1.2 Summarizing and graphing: 

**I will summarize one tibble and plot one graph for each question.** 

###First Question: How many trees were planted per neighborhood every year?

**Summarizing:** 

I am going to create a new tibble with three columns named tree_neighbourhood which contains neighbourhood name, year of tree planted and number of trees planted that year by using dplyr verbs. I will use this tibble later to plot the graph. 

```{r}
tree_neighbourhood<-vancouver_trees %>%
  group_by(neighbourhood_name, plant_year = year(date_planted)) %>%
  summarise(number_of_trees = n()) %>%
  drop_na()
print(tree_neighbourhood)
```

Now, the tibble is ready to use. I am going to plot the histogram and path of the number of trees respect to the year that the trees were planted and separate them into each neighbourhood by using dplyr and ggplot2. Note: I adjust the fill and alpha transparency for the histogram in order to see the path better. I set the range of number of trees to 0 and 500 so it is easier to see the relationship. 

**Graphing:** 

```{r fig.height = 15, fig.width = 12}
tree_neighbourhood %>%
  group_by(neighbourhood_name) %>%
  drop_na() %>% #drop null columns
  ggplot(aes(x=plant_year, y= number_of_trees)) +
  facet_wrap(~neighbourhood_name,ncol=3)+
  geom_histogram(stat="identity", aes(alpha=0.3),fill="cornflowerblue") +
  coord_cartesian(ylim =  c(0,500))+
  geom_path()+geom_point(size=1,color="blue") +
  xlab("plant year")+ylab("number of trees")+
  theme_light()
```




###Second Question: What's the relationship between the height of the tree and the age of the tree for each neighbourhood?

I want to find the relationship between height of the tree and the age of the tree. I am going to create a tibble named height_sub which contains the name of neighbourhood, age of tree, the mean/minimum/maximum/median/standard deviation of the height, then I arranged the tibble by the age of the tree. The height range is is between 0-10 for every 10 feet (e.g., 0 = 0-10 ft, 1 = 10-20 ft, 2 = 20-30 ft, and 10 = 100+ ft). For example, the mean of the height is 1 means the mean of the height is between 10-20ft. 

**Summarizing:**

```{r}
height_sub<-vancouver_trees %>%
  group_by(neighbourhood_name, age_of_tree = 2021 - year(date_planted)) %>%
  summarise(height_mean=round(mean(height_range_id, na.rm=TRUE), 2),
            height_min=min(height_range_id, na.rm=TRUE), 
            height_max=max(height_range_id, na.rm=TRUE), 
            height_median=median(height_range_id, na.rm=TRUE),
            height_sd=sd(height_range_id, na.rm=TRUE))%>%
  arrange(age_of_tree)
print(height_sub)

```
I am going to plot the mean of the height with respect to the age of tree for each neighbourhood using the geom_point, geom_line, and facet_wrap.

**Graphing:** 

```{r fig.height = 15, fig.width = 13}
height_sub %>%
  group_by(neighbourhood_name) %>%
  drop_na() %>% #drop null columns
  ggplot(aes(x=age_of_tree, y= height_mean)) +
  geom_point(size=1.5, color="purple") + 
  geom_line(aes(alpha=0.4),colour = "blue")+
  facet_wrap(~neighbourhood_name, ncol = 4)+
  ylab("mean of the height") + xlab("age of tree")
  
```



###Third Question: Does the diameter different for each species? 

**Summarizing:** 
  
To find out if the diameter is different for each species, I am going to create a tibble named tree_diameter which contains species name, mean/minimum/maximum/median/standard deviation of the diameter for each species, and the number of trees of each species by using summarise(). Then I arranged the tibble using the mean of diameter.  

```{r}
tree_diameter <- vancouver_trees %>%
  select(diameter, species_name) %>%
  group_by(species_name) %>%
  summarise(diameter_mean=mean(diameter, na.rm=TRUE), diameter_min=min(diameter, na.rm=TRUE), diameter_max=max(diameter, na.rm=TRUE), diameter_median=median(diameter, na.rm=TRUE), diameter_sd=sd(diameter, na.rm=TRUE), n=n()) %>%
  arrange(across(diameter_mean))
print(tree_diameter)
```

**Graphing:** 

I am going to use the tibble above to plot the point and line of mean and median of the trees with respect to each species. Since there are 283 species, it is impossible to plot all of them. I want to plot only those species that have more than 2000 trees so we have enough samples to analyze. 

```{r fig.width=10}
tree_diameter %>%
  filter(n >2000) %>%
  # Filter the species that have more than 2000 trees
  ggplot(aes(species_name)) + 
  geom_line(aes(y = diameter_mean, colour = "mean",group=1)) + 
  geom_point(aes( y = diameter_mean)) +
  geom_line(aes(y = diameter_median, colour = "median", group=1))+
  geom_point(aes( y = diameter_median)) +
  coord_cartesian(ylim =  c(0,40))+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```




###Fourth Question: Does the tree diameter different in each neighbourhood with respect to age of the tree?

**Summarizing:** 

For the summarizing part, I am going to group the data by neighbourhood name and the year plant the tree, and then use summarise() to find the mean/minimum/maximum/median/standard deviation of the diameter for neighbourhood. It is arranged by the year, so it is easier to see if the diameter is different. 


```{r}
vancouver_trees %>%
  group_by(neighbourhood_name, age_of_tree=2021 - year(date_planted))%>%
  summarise( diameter_mean=mean(diameter, na.rm=TRUE), diameter_min=min(diameter, na.rm=TRUE), diameter_max=max(diameter, na.rm=TRUE), diameter_median=median(diameter, na.rm=TRUE), diameter_sd=sd(diameter, na.rm=TRUE), n=n()) %>%
  arrange(across(age_of_tree))
```
**Graphing:** 

For the graph, I am going 

```{r fig.height = 15, fig.width=12}
vancouver_trees %>%
  select(neighbourhood_name, date_planted, diameter) %>%
  drop_na() %>%
  mutate(age_of_tree = 2021 - year(date_planted)) %>%
  mutate(age_range = case_when(age_of_tree<6 ~ "0 to 6",
                               age_of_tree<11 ~ "6 to 11",
                               age_of_tree<16 ~ "11 to 16",
                               age_of_tree<21 ~ "16 to 21",
                               age_of_tree<26 ~ "21 to 26",
                               TRUE ~ "26-")) %>%
  group_by(neighbourhood_name) %>%
  filter(n() > 3000 ) %>%
  ggplot(aes(x=age_range,y= diameter))+
  scale_x_discrete(limits = c("0 to 6","6 to 11", "11 to 16","16 to 21", "21 to 26", "26-")) +
  geom_boxplot(aes(alpha=0.7)) +
  geom_jitter(size=0.7, aes(alpha=0.6)) +
  facet_wrap(~neighbourhood_name, ncol = 5) +  
  coord_cartesian(ylim =  c(0, 25))+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```



### 1.3 

The following are my analysis for each questions:

1. *How many trees were planted per neighborhood every year?* 
The number of trees were planted per neighbourhood are quite different every year. And it does not have a clear pattern how they decide how many trees they are going to plant. I want to analyze this question further into more detail. I decided to pick the Kerrisdale neibourhood to analyze how many trees were plant every year/season/month. 

2. *What's the relationship between the height of the tree and the age of the tree for each neighbourhood? * 
The graph shows a positive relationship between the mean of the height and the age of the tree in each neighbourhood.   

3. *Does the diameter of trees different for each species?*

The mean and median of the diameter for each species are quite different. I decide to pick the AMERICANO species to analyze the relationship of diameter and the age of the tree for this specific species. 

4. *Does the diameter of trees different in each neighbourhood* 
The result is yes. As the age become larger, the diameter tends to have bigger difference. 


# Task 2: Tidy the data 

The definition of *tidy* data:

- Each row is an **observation**
- Each column is a **variable**
- Each cell is a **value**

*Tidy'ing* data is sometimes necessary because it can simplify computation. Other times it can be nice to organize data so that it can be easier to understand when read manually. 



### 2.1

I selected the following 8 columns and store them in to the tibble named vancouver_trees_sub. I will use this tibble for later tasks. 

```{r}
vancouver_trees_sub <- vancouver_trees %>%
  select(genus_name,species_name,cultivar_name,common_name,neighbourhood_name,height_range_id, diameter,date_planted)
print(vancouver_trees_sub)
```

genus_name: This column contains the different genus name of each tree. 

species_name: This column contains the different species name of each tree. 

cultivar_name: This column contains the cultivar  name of each tree. 

common_name: This column contains the common  name of each tree.

neighbourhood_name: This column contains the name of neighbourhood where the tree is located. 

height_range_id: The height range is is between 0-10 for every 10 feet (e.g., 0 = 0-10 ft, 1 = 10-20 ft, 2 = 20-30 ft, and 10 = 100+ ft).

diameter: The diameter of the tree in inches.

date_planted: The date of planting in YYYYMMDD format. 

**This tibble is tidy. Each column is a distinct variable and it is tidy. Each cell of the row contains a value respect to this variable. Each row is an unique observation from the dataset.** 



### 2.2 

Since the data is tidy. I am  going to untidy the data first. Then I will tidy it back to the original tibble.`

**Untidy:**

I am going to untidy it by using the pivot_longer() to combine the columns(genus_name, species_name,cultivar_name, common_name) together. This is untidy because for my research questions, I am analyze the relationship of diameter with respect to species. If these columns are combined together, I will need to use pivot_longer to tidy it first to do the analysis. 

```{r}
vancouver_trees_sub1 <- vancouver_trees_sub %>%
  pivot_longer(cols = c("genus_name":"common_name"), 
              names_to = "name_class", values_to = "name_of_tree")
print(vancouver_trees_sub1)
```

**Tidy:**

The columns(genus_name, species_name,cultivar_name, common_name) are combined into two columns(name_class and name_of_tree). This data is untidy right now, so I am going to use pivot_wider to make it tidy again.   
```{r}
vancouver_trees_sub2 <- vancouver_trees_sub1 %>%
  pivot_wider( names_from = name_class,
                values_from = name_of_tree) %>%
  unnest()
print(vancouver_trees_sub2)
```



### 2.3 

I want to analyze the following two questions in milestone 3:

I modified my questions to be more specific from the original questions which are too broad. So I pick one specific species and one specific neighbourhood to analyze.   

1. What's relationship of diameter with respect to age of the tree for AMERICANO species. 

2. How many trees were planted every year in Kerrisdale neighbourhood? How about every season? Every month? 


Next, I am going to modify the data to make it be more appropriate to answer the above questions in milestone 3. To answer the 2 questions, I will select the following columns: species_name, diameter, date_planted, neighbourhood_name. Then I am going to filter the rows that either the species is AMERICANA or the neighbourhood_name is KERRISDALE. Also, I am going to add three columns which are the age of the tree in years, month and the season that the tree was planted. Since I will no longer need the column of date_planted, I will drop this column at the end. 

```{r}
vancouver_trees %>%
  select(species_name,  diameter, date_planted, neighbourhood_name) %>%
  filter(species_name == "AMERICANA" | neighbourhood_name == "KERRISDALE") %>%
  drop_na() %>%
  mutate(age_of_tree = round(as.numeric((ymd(20211004) - ymd(date_planted))/365),digits=2)) %>%
  mutate(plant_month = month(date_planted)) %>%
  mutate(plant_season = case_when(plant_month < 3 ~ "Winter",
                                  plant_month < 6 ~ "Spring",
                                  plant_month < 9 ~ "Summer",
                                  plant_month < 12 ~ "Fall",
                                  TRUE ~ "Winter")) %>%
  select(-date_planted)
```


