---
title: "Mini data analysis - vancouver_trees dataset"
author: "Yuwei Yang"
date: "2021/10/5"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
```

Please install the following three packages before running the code. The package datateachr contains 7 datasets and we will use 4 of them. The package tidyverse contains the package dplyr and we will use it to do data wrangling. The package ggplot2 is used to do visualization in this analysis.

```{r}
library(datateachr)
library(tidyverse)
library(ggplot2)
```

# Task 1: Choose the dataset

## 1.1 Choose 4 datasets from datateachr packages:
  1: steam_games: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.
  
  2: flow_sample: Acquired courtesy of The Government of Canada’s Historical Hydrometric Database. It currently has 218 rows and 7 columns.
  
  3: parking_meters: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 10032 rows and 22 columns.
  
  4: cancer_sample: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

## 1.2 Explore datasets: 

In this part, I will check the class of the data using class(), and the number of rows and columns, and the name of each column with some useful information using glimse(). You may also use ?dataset to get the description and other relative information for this dataset.

### The following is the information for steam_games data:

```{r}
class(steam_games)
glimpse(steam_games)
#?steam_games
```

### The following is the information  for flow_sample data:

```{r}
class(flow_sample)
glimpse(flow_sample)
```


### The following is the information or vancouver_trees data:

```{r}
class(vancouver_trees)
glimpse(vancouver_trees)
```


### The following is the information  for cancer_sample data:

```{r}
class(cancer_sample)
glimpse(cancer_sample)
```

# 1.3 Select dataset:

The following are the questions I have for each dataset: 

*steam_games:* 

 1. Which developer and publisher receive the most achievements?
 
 2. Does the achievements of the game relate to the reviews(both recent and all reviews)?

*flow_sample:*

 1. What's the relationship of the flow rate with respect to year? 
 
 2. What's the busiest day every week?

*Vancouver_trees:*

 1. What's the distribution of the number of trees with respect to different height range?   
 
 2. What's the relationship between tree diameter and the age of the tree.  
 
 3. Will this relationship be different for each species? 

*Cancer_sample:* 

1. For different diagnoses, will the parameters be different? 


I decide to analyze vancouver_trees because I am currently living in Vancouver, so I am very interested in the data of this city. Also, this dataset seems to have some interesting relationships between each variable. For example, I am interested in whether the diameter of a tree will be affected by other factors such as species.   


# Task 2: Explore dataset - vancouver_trees

## The following three questions are the goal of this analysis:
  1. What's the distribution of the number of trees with respect to different height range?
  2. What's the relationship between tree diameter and the age of this tree.
  3. Will this relationship be different of each species? 


## 1. The first question I am interested in is what the distribution of the number of trees with respect to different height range is. I am going to use ggplot2 to plot the distribution of the number of trees using different height range id.  The height range is is between 0-10 for every 10 feet (e.g., 0 = 0-10 ft, 1 = 10-20 ft, 2 = 20-30 ft, and 10 = 100+ ft).

```{r}
ggplot(vancouver_trees, aes(height_range_id)) +
  geom_histogram(bins=11, binwidth=1, fill ="pink", alpha=0.8)+
  xlab("Tree Height Range ID")+ylab("Number of Trees")+
  theme_classic()
```

From the histogram above, we can see that most of the trees have a height between 10 and 30 ft.

## 2. The second question I want to analyze is the relationship between tree diameter and the age of the tree. I am going to creat a new tibble named vancouver_trees_sub which contains species name and diameter. Then, I will add a new column named tree_age which is the number of years between the tree were planted and the date that this dataset was last updated "2021-10-04"(round to 2 decimal places) by using dplyr verbs. I will use this tibble later to find the relationship between the age of the tree and the height of the tree. 

```{r}
vancouver_trees_sub <- vancouver_trees %>%
  # Calling difftime() function
  mutate(tree_age = round(as.numeric(difftime("2021-10-04", date_planted, units="days"))/365, 2)) %>%
  # add tree_age column to vancouver_trees tibble
  select(species_name, diameter,tree_age)

head(vancouver_trees_sub)
```

## 3. Next, I am going to find out if there is a relationship between the age of the tree and the diameter of the tree. I will use ggplot2 package to plot the jitterplot. I set the range of diameter to 0 and 50 because I want to get rid of the outliers in the graph, so it is easier to see the relationship. 

```{r}
ggplot(vancouver_trees_sub, aes(tree_age, diameter))+
  geom_jitter(outlier.shape = NA, aes(alpha=0.4, colour=tree_age))+
  coord_cartesian(ylim =  c(0, 50))+
  xlab("Age of tree")+ylab("Diameter")
```


## 4. As we can see, from the jitterplot, it seems that the diameter of trees tends to be larger as the age getting larger. So, next, I want to analyze this relationship in detail by separating the trees into different species and look at the relationship within each species. I will use both dplyr and ggplot2 in this part. Also, I want to plot only those species that have more than 5000 trees so we have enough samples to analyze. You may be noticed that I set the range of the diameter to 0 and 35, and this is because I want to get rid of the most outliers for these species. 


```{r}
vancouver_trees_sub %>%
  group_by(species_name) %>%
  filter(n() >5000) %>%
  # Filter the species that have more than 5000 trees
  ggplot(aes(tree_age, diameter))+ 
  geom_point(aes(alpha=0.4, colour=tree_age),size=1)+  
  facet_wrap(~species_name)+
  coord_cartesian(ylim =  c(0,35))+
  xlab("Age of tree")+ylab("Diameter")

```

Again, the diameter of the tree seems to have a positive relationship with the age of the tree within each species. And some species tend to have a larger diameter than others such as Serrulata tend to have a larger diameter when their age getting older.  



# Task 3: Research Questions
  1. Will the root barrier affect the diameter of the tree?
  2. What's the relationship between diameter and plant_area?
  3. What's the relationship between the height and the age of the tree?
  4. How many trees are there in each neighborhood? 

These are the four questions I want to analyze. First of all, I want to further explore the relationship of the diameter with other factors, so I pick root barrier and plant_are. I am not sure if there are other factors related to diameter. I would like to have suggestions on this. Also, I think height might relate to the age of the tree, but with height range, it might cause some problems. 

