---
title: "Mini-Data-Analysis 3"
author: "Yuwei"
date: "2021/10/24"
output: github_document
---


# Goal of this project:

+ Manipulating special data types in R: factors and/or dates and times.

+ Fitting a model object to your data, and extract a result.

+ Reading and writing data as separate files.

# Setup

Please install the following four packages before running the code. The package datateachr contains 7 datasets and I will use vancouver_trees dataset for this analysis. The package tidyverse contains the package dplyr, forcats and tidyr, and I will use it to do data wrangling and tidy data. The package ggplot2 is used to do visualization in this analysis. The package lubridate provides tools to parse and manipulate dates. 


```{r, message = FALSE}
library(datateachr) #contain the dataset we will be using
library(tidyverse) 
library(ggplot2)
library(lubridate)
```

For this Mini Data Analysis 3, I will be using the vancouver_trees dataset. This analysis contains three exercises: 1. Special Data types 2. Modelling 3. Reading and writing data

# Research questions:

1. **What's relationship of diameter with respect to age of the tree for AMERICANO species.** 

2. **How many trees were planted every year in Kerrisdale neighbourhood? How about every season? Every month?** 


# Exercise 1: Special Data Types

In this exercise, I will modify the plot I made in milestone 3. The code and the plot is placed below: 


### Plot from milestone 2: 

I divided the age of the trees into 5 groups( 0 to 6, 6 to 11, 11 to 16, 16 to 21, 21 to 26, and 26 +), and use filter to select neighbourhood that have more than 4000 trees because there are a lot of neighbourhoods and it is hard to plot all of them. I am going to plot the boxplot and jitterplot of the diameter for each age group in each neighbourhood. 

```{r}
vancouver_trees %>%
  select(neighbourhood_name, date_planted, diameter) %>%
  drop_na() %>% # drop all the rows that contain NULL
  mutate(age_of_tree = 2021 - year(date_planted)) %>%
  #add the column of age of the tree
  mutate(age_range = case_when(age_of_tree<6 ~ "0 to 6",
                               age_of_tree<11 ~ "6 to 11",
                               age_of_tree<16 ~ "11 to 16",
                               age_of_tree<21 ~ "16 to 21",
                               age_of_tree<26 ~ "21 to 26",
                               TRUE ~ "26+")) %>%
  #assign the tree into each age group
  group_by(neighbourhood_name) %>%
  filter(n() > 4000 ) %>%
  # Filter the neighbourhood that have more than 2000 trees
  ggplot(aes(x=age_range,y= diameter))+
  geom_jitter(size=0.7, colour = "pink") +
  geom_boxplot(aes(alpha = 0.3)) +
  facet_grid(~neighbourhood_name) +  
  coord_cartesian(ylim =  c(0, 25))+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

Now, we have the plot from previous milestone. I am going to choose two tasks from the below three tasks and modify this plot by the instructions. 


1. Produce a new plot that reorders a factor in your original plot, using the `forcats` package (3 points). Then, in a sentence or two, briefly explain why you chose this ordering (1 point here for demonstrating understanding of the reordering, and 1 point for demonstrating some justification for the reordering, which could be subtle or speculative.)

2. Produce a new plot that groups some factor levels together into an "other" category (or something similar), using the `forcats` package (3 points). Then, in a sentence or two, briefly explain why you chose this grouping (1 point here for demonstrating understanding of the grouping, and 1 point for demonstrating some justification for the grouping, which could be subtle or speculative.)

3. If your data has some sort of time-based column like a date (but something more granular than just a year):
    1. Make a new column that uses a function from the `lubridate` or `tsibble` package to modify your original time-based column. (3 points)
        - Note that you might first have to _make_ a time-based column using a function like `ymd()`, but this doesn't count.
        - Examples of something you might do here: extract the day of the year from a date, or extract the weekday, or let 24 hours elapse on your dates.
    2. Then, in a sentence or two, explain how your new column might be useful in exploring a research question. (1 point for demonstrating understanding of the function you used, and 1 point for your justification, which could be subtle or speculative).
        - For example, you could say something like "Investigating the day of the week might be insightful because penguins don't work on weekends, and so may respond differently".
But first, tasks 1 and 2 below ask you to modify a plot you made in a previous milestone. The plot you choose should involve plotting across at least three groups (whether by facetting, or using an aesthetic like colour). Place this plot below (you're allowed to modify the plot if you'd like). If you don't have such a plot, you'll need to make one. Place the code for your plot below.


**Task Number**: 3

In the previous analysis, I calculated the age of the tree using the current year subtract the plant year. But I ignored the time difference between the plant month and date. So the age I calculated was not very accurate. I want to add this difference to make the age more accurate. 

Instead of using the previous way to calculate the age of the trees, I will calculate the time difference between the plant date and the and the date that this dataset was last updated "2021-10-04"(round to 2 decimal places) by using the lubridate package. First, I am going to add a column named time_diff which is the time difference in days. Then I am going add a new column called age_of_tree which are the time difference converted in years. 

Then, the trees are divided 5 age groups( 0 to 6, 6 to 11, 11 to 16, 16 to 21, 21 to 26, and 26 +) by using the age of the trees we just calculated. This new tibble is named as tree_age_range and I am going to use this tibble to find the distribution of trees by different age groups. 

```{r}
tree_age_range <- vancouver_trees %>%
  select(neighbourhood_name, date_planted, diameter) %>%
  drop_na() %>% # drop all the rows that contain NULL
  mutate(time_diff = ymd(20211004)-date_planted) %>%
  #add the column of time difference in day
  mutate(age_of_tree = as.numeric(round(time_diff/365, digits=2))) %>%
  mutate(age_range = case_when(age_of_tree<=6 ~ "0 to 6",
                               age_of_tree<=11 ~ "6 to 11",
                               age_of_tree<=16 ~ "11 to 16",
                               age_of_tree<=21 ~ "16 to 21",
                               age_of_tree<=26 ~ "21 to 26",
                               TRUE ~ "26+")) %>%
  select(diameter, age_range)
  
print(tree_age_range)
```



**Task Number**: 1

In the graph below, the x-axis age_range is out of order. So I am going to use fct_relevel() to reorder the x axis. 

```{r}
tree_age_range %>%
  ggplot(aes(age_range))+
  geom_bar()
```
The graph below is the updated graph after reordering x axis. After relevel, we can see the graph makes more sense now. The graphs shows that most trees are between age 21-26. 

```{r}
tree_age_range %>%
  mutate(age_range = fct_relevel(age_range, c("0 to 6","6 to 11", "11 to 16","16 to 21", "21 to 26", "26-"))) %>%
  ggplot(aes(age_range))+
  geom_bar()
```


    

# Exercise 2: Modelling

## 2.0 

The following is the research question and variable of interest I am going to analyze in this exercise:

**Research Question**: What's relationship of diameter with respect to age of the tree for AMERICANO species. 


**Variable of interest**: diameter


## 2.1 

I am going to make a tibble named Americana_tree which contains the diameter and date_planted for AMERICANA species. Then I am going to add a new column called age_of_tree. Next, I am going to fit a linear model on diameter with respect to age of tree and store the model as model1. 

The original plot has some big outliers. 
 

```{r}
Americana_tree <- vancouver_trees %>%
  select(species_name,  diameter, date_planted) %>%
  filter(species_name == "AMERICANA" ) %>%
  drop_na()%>%
  mutate(age_of_tree = round(as.numeric((ymd(20211004) - ymd(date_planted))/365),digits=2))

model1 <- lm(diameter~age_of_tree, data = Americana_tree)

print(model1)
```

To see the accuracy of model1, I am going to plot the data as points and model1 as regression line. It is obvious that there are some extreme outliers which affect the accuracy of our model. So next, I am going to remove the outliers. 

```{r}
Americana_tree %>%
  ggplot(aes(age_of_tree, diameter)) +
  geom_point() +
  geom_abline(slope=0.4027,intercept=-0.7912)
```

Removing outliers, I am going to store all the outliers in the variable called outliers. And then I am going to remove these outliers from the Americana_tree data and name the new tibble as Americana_tree_2. Next, I am going to fit a new linear model using the new dataset without outliers. 

```{r}
outliers <- boxplot(Americana_tree$diameter, plot=FALSE)$out
Americana_tree_2 <- Americana_tree[-which(Americana_tree$diameter %in% outliers),]
#get rid of outliers
model2 <- lm(diameter~age_of_tree, data = Americana_tree_2)
print(model2)
```

Below is the graph of the data and new model:

```{r}
Americana_tree_2 %>%
  ggplot(aes(age_of_tree, diameter)) +
  geom_point() +
  geom_abline(slope=0.3765,intercept=--0.8627)
```

To test if the accuracy of the model was increased by removing outliers, I summarized the two models. And it is clear that R squared is increased more than double, so I think the model was improved a lot by removing outliers. 

```{r}
summary(model1)
summary(model2)
```


<!----------------------------------------------------------------------------->

## 2.2 

I am going to produce a new tibble named model2_fit which includes the predictions on diameter and the residule using the model2 I fitted previously using broom::augment().
  

```{r}
model2_fit<- broom::augment(model2) %>% 
  rename(prediction_of_diameter = .fitted, residule = .resid) %>%
  select(diameter, age_of_tree, prediction_of_diameter, residule)

print(model2_fit)
```



# Exercise 3: Reading and writing data

For this exercise, I am going to create a new folder called output in the top level of the repository. 

## 3.1 (5 points)


Take a summary table that you made from Milestone 2 (Exercise 1.2), and write it as a csv file in your `output` folder. Use the `here::here()` function.

- **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
- **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file. 
  
<!-------------------------- Start your work below ---------------------------->

```{r}

```

<!----------------------------------------------------------------------------->

## 3.2 (5 points)

Write your model object from Exercise 2 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`. 

- The same robustness and reproducibility criteria as in 3.1 apply here.

<!-------------------------- Start your work below ---------------------------->


<!----------------------------------------------------------------------------->

# Tidy Repository

Now that this is your last milestone, your entire project repository should be organized. Here are the criteria we're looking for.

## Main README (3 points)

There should be a file named `README.md` at the top level of your repository. Its contents should automatically appear when you visit the repository on GitHub. 

Minimum contents of the README file:

- In a sentence or two, explains what this repository is, so that future-you or someone else stumbling on your repository can be oriented to the repository.
- In a sentence or two (or more??), briefly explains how to engage with the repository. You can assume the person reading knows the material from STAT 545A. Basically, if a visitor to your repository wants to explore your project, what should they know?

Once you get in the habit of making README files, and seeing more README files in other projects, you'll wonder how you ever got by without them! They are tremendously helpful.

## File and Folder structure (3 points)

You should have at least four folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else. 

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (2 points)

All output is recent and relevant: 

- All Rmd files have been `knit`ted to their output, and all data files saved from Exercise 3 above appear in the `output` folder. 
- All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
- There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted. 

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant. 

PS: there's a way where you can run all project code using a single command, instead of clicking "knit" three times. More on this in STAT 545B!

## Error-free code (1 point)

This Milestone 3 document knits error-free. (We've already graded this aspect for Milestone 1 and 2)

## Tagged release (1 point)

You've tagged a release for Milestone 3. (We've already graded this aspect for Milestone 1 and 2)